### 2011-06-05 docs/pubs/EIM-2011/framework.tex
Trivial spelling and grammar fixes.
added more to annotation and materialization based on matt's suggestions

\mypara{Semantic Annotations.} A semantic annotation consists of two
parts: (i) a ``configuration'' of the observation model containing the
specific entities, characteristics, observations, measurements, and so
on (drawn from one or more domain ontologies) that appropriately
capture the semantics of the data set; and (ii) a mapping between the
attributes in the data set to specific measurements defined in the
model configuration. \figref{fig:kelp-mass-model} shows a high-level
example of an annotation defined for a simple Kelp sampling data
set. Here, the data set consists of five attributes (bottom of
\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
measurement type (where only the characteristic of each measurement
type is shown), and measurement types are organized into observations
of specific Kelp entities (shown of type ``Macrocystis''), temporal
points (denoted by date-times), and spatial locations (given as site
names). Each measurement associated with a Kelp observation is assumed
to have occurred within the site and during the given time as
specified by the context relationships.  Semantic annotations provide
a formal description of attribute semantics, whereas in many
commonly-used metadata formats for describing data sets,  only
informal text-based descriptions of data attributes are permitted.  In
the case of EML, there is some overlap between these two
mechanisms---particularly with respect to measurement
standards---however, semantic annotations extend this approach by
providing a general mechanism to formally associate concepts drawn
from domain ontologies to attributes. In the Semtools framework, we
employ an XML serialization syntax for semantic annotations that is
compatible with EML but that is stored separately from the EML
documentation of a data set (allowing, e.g., annotations to be used
independently of EML or with other metadata standards if needed). In addition, semantic annotations can
be used to ``materialize'' a given data set into a set of triples
conforming to the model configuration given in the annotation.
Materializing a data set to a set of OBOE triples in this way provides
a more uniform structural representation that can make a number of
integration tasks easier (described further below).

@@

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either client or server deployments).
The Semantic Mediation API uses both the OWL API \cite{owlapi} for ontology
management services including RDF/OWL parsing, serialization, and
simple class and property exploration as well as the Pellet
description-logic reasoner \cite{pellet} for classification and exposing
inferred axioms in source ontologies. The inference services exposed
through the Semantic Mediation API are used in both discovery and
integration described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

>>>>

\mypara{Semantic Annotations.} A semantic annotation consists of two
parts: (i) a ``configuration'' of the observation model containing the
specific entities, characteristics, observations, measurements, and so
on (drawn from one or more domain ontologies) that appropriately
capture the semantics of the data set; and (ii) a mapping between the
attributes in the data set to specific measurements defined in the
model configuration. \figref{fig:kelp-mass-model} shows a high-level
example of an annotation defined for a simple Kelp sampling data
set. Here, the data set consists of five attributes (bottom of
\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
measurement type (where only the characteristic of each measurement
type is shown), and measurement types are organized into observations
of specific Kelp entities (shown of type ``Macrocystis''), temporal
points (denoted by date-times), and spatial locations (given as site
names). Each measurement associated with a Kelp observation is assumed
to have occurred within the site and during the given time as
specified by the context relationships.

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp
  entities}'' and ``\emph{find all data sets containing Mass
  measurements of Kelp}''.\footnote{Note that queries also can include
  contextual constraints, e.g., Mass measurements of Kelp within a
  specific spatial or temporal context.} Both of these queries would
return the example data set in \figref{fig:kelp-mass-model} since the
attribute {\tt DRY} is linked to a DryMass measurement of observations
of Macrocystis (where DryMass is defined as a special kind, or
\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
Kelp). Using semantic annotations in this way can help to increase
both query recall and precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched), and the precision of the result can be improved since
queries can specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
Annotations also help facilitate integration by allowing tools to
align data-set attributes based on their declared measurement and
observation types.

In general, semantic annotations provide a formal description of
attribute semantics, whereas in many commonly-used metadata formats
for describing data sets, only informal text-based descriptions of
data attributes are permitted.  In the case of EML, there is some
overlap between these two mechanisms---particularly with respect to
measurement standards---however, semantic annotations extend this
approach by providing a general mechanism to formally associate
concepts drawn from domain ontologies to attributes. In the Semtools
framework, we employ an XML serialization syntax for semantic
annotations that is compatible with EML but that is stored separately
from the EML documentation of a data set (allowing, e.g., annotations
to be used independently of EML or with other metadata standards if
needed). In addition, semantic annotations can be used to
``materialize'' a given data set into a set of triples conforming to
the model configuration given in the annotation. In other words, a
tabular data set such as the one shown in \figref{fig:kelp-mass-model}
can automatically be converted into a corresponding collection of
observation and measurement instances. This in turn enables a simple
form of structural integration, where instead of having a large number
of different tabular data structures, all data is represented using
the standard set of structures defined by the OBOE model (see
\figref{fig:oboe}). Thus, materializing a data set in this way
provides a more uniform structural representation that can make a
number of discovery and integration tasks easier. For instance,
materialization can be used to increase query expressivity by allowing
searches of the form ``\emph{find all data sets containing Mass
  measurements of Kelp with values less than 1 Kilogram}'', which in
our example can be answered by generating (i.e., materializing) the
measurements associated with the {\tt DRY} attribute in the data set
of \figref{fig:kelp-mass-model}.

@@

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either client or server
deployments).  The Semantic Mediation API uses both the OWL API
\cite{owlapi} for ontology management services including RDF/OWL
parsing, serialization, and simple class and property exploration as
well as the Pellet description-logic reasoner \cite{pellet} for
classification and exposing inferred axioms in source ontologies. The
inference services exposed through the Semantic Mediation API are used
in both discovery and integration features described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.
